Dynamics of Laterally Propagating Flames in X-ray Bursts. II. Realistic Burning \& Rotation

### 2020-12-30 xrb2/paper.tex
added citation to Lupe's thesis
Migrate discussion of Rossby length to Section 3.1
The new additions to Section 3.1 are in blue.

We run three models ({\tt F250}, {\tt F500}, and {\tt F1000}) with the same initial model in terms of temperature but differing rotation rates. The {\tt F500} and {\tt F1000} runs both qualitatively resemble the flame structure in \citet{flame_wave1} --- a laterally propagating flame that is lifted off of the bottom of the fuel layer --- but they differ in their burning structures. Figures~\ref{fig:time_series_500} and \ref{fig:time_series_1000} show time series of the mean molecular weight, $\bar{A}$, for the {\tt F500} and {\tt F1000} runs.  Compared to those in \cite{flame_wave1}, ashes behind the flame do not reach as high atomic weights. This is not surprising, since those previous runs artificially boosted the reaction rates.  Comparing these two new runs, the burning is much more evolved for the higher rotation rate, and the ash is actually able to move ahead of the flame front (visible in the Figure  \ref{fig:time_series_1000} 100 ms snapshot). We believe that this is because the increased rotation better confines the initial perturbation and subsequent expansion from the burning, increasing the temperature and density in the flame front such that the reaction rate increases, which allows the reactions to progress further. In contrast to {\tt F500} and {\tt F1000}, the lowest rotation run --- {\tt F250} --- failed to ignite. The lack of ignition for {\tt F250} also aligns with the reasoning given above, with the lower rotation in this case potentially leading to insufficient confinement such that the temperature and density required for ignition is not achieved. In this scenario, another source of confinement (e.g.\ magnetic fields) would need to take over at lower rotation rates to allow a burst to occur, at least for the initial model used here.

@@

We found in \citet{flame_wave1} that the flame speed, $s$, obeys the following relation:
\begin{equation}
	\label{eqn:flame_speed_scaling}
	s \appropto L_R (\enucdot)^{\half} \propto \frac{(\enucdot)^{\half}}{\Omega},
\end{equation}
where $\enucdot$ is the specific nuclear energy generation rate and $L_R$ is the Rossby length. 
This was consistent with the results of \citet{cavecchi:2013}, who noted that the flame speeds in 
their simulations scaled with the Rossby length. As seen in Figure~\ref{fig:flame_speeds_1}, there 
is no obvious inverse scaling of the flame speed with rotation rate in this set of runs. We 
observed earlier, however, that nuclear reactions progress more quickly at higher rotation 
rate. This results in a higher $\enucdot$ --- up to four to six times higher near the burning front 
after the flame ignites --- which may counteract the reduction in flame speed from the increased 
Coriolis confinement. The initial position of $\tt{F1000}$ does appear to be reduced compared to 
the lower rotation $\tt{F500}$ run. From Equation \ref{eqn:flame_speed_scaling}, increasing the 
rotation rate from 500 Hz to 1000 Hz should decrease $L_R$ by a factor of two, and the initial 
position of the flame should be reduced by a similar factor (again, due to greater confinement from 
the Coriolis force). $L_R$ is not, however, a perfect metric for the flame lengthscale. The scaling 
described in Equation \ref{eqn:flame_speed_scaling} seems to be more accurate at higher rotation 
rates (1000 Hz and up), based on \citet{flame_wave1}. We also observe that $\tt{F500}$ accelerates 
faster than $\tt{F1000}$, which appears to experience a small deceleration at early times. 
\MarginPar{KE: This is speculation on my part -- does someone else have clear idea why?} This 
disparity may be a direct result of the difference in Coriolis force. It could also be due to the 
500 Hz flame taking more time to develop fully and reach higher burning rates, resulting in an 
$\enucdot$ increase over time.

@@

In future simulations, we would like to improve and expand our simulations in order to better understand the processes at play and to include more physics. This includes adding tracer particles to the simulations so we can monitor the fluid motion; extending our simulations to 3D, which would hopefully alleviate some of the boundary effects we have observed in these simulations but will require significant computational resource; and exploring the resolution of our simulations more so that we can ensure that we have resolved all the necessary physical processes. We would also like to model H/He flames, \MarginPar{someone please describe here why these would be interesting as I can't remember what makes them different}. We recently added an MHD solver to \castro; this will allow us to explore the effects of magnetic fields on flame propagation in XRBs.

>>>>

We run three models ({\tt F250}, {\tt F500}, and {\tt F1000}) with the same initial model in terms 
of temperature but differing rotation rates. {\color{blue}We saw in \citet{flame_wave1} that 
increasing the 
neutron star rotation rate reduces the horizontal lengthscale of the flame. An estimate of 
this lengthscale is given by the Rossby radius of deformation, $L_R$. The Rossby radius may be 
thought of as the scale over which the balance between the Coriolis force and horizontal pressure 
gradient becomes important. It is approximately:
\begin{equation}
	\label{eqn:rossby}
	L_R \approx \frac{\sqrt{g H_0}}{\Omega},
\end{equation}
where $g$ is gravitational acceleration, $H_0$ is the atmospheric scale height, and $\Omega$ is the 
neutron star rotation rate. Comparing Figure \ref{fig:time_series_enuc_500} and Figure 
\ref{fig:time_series_enuc_1000}, the horizontal extent of the $\tt{F1000}$ flame does appear to be 
reduced compared to the lower rotation $\tt{F500}$ run. From Equation \ref{eqn:rossby}, increasing 
the rotation rate from 500 Hz to 1000 Hz should decrease $L_R$ by a factor of two, and the greater 
confinement from the Coriolis force should reduce the horizontal extent of the flame by a similar 
factor. The Rossby radius is not a perfect measure of this horizontal lengthscale, however, and 
here the difference is not a factor of two. The simulations seem to follow the theoretical
scaling described in Equation \ref{eqn:rossby} more closely at higher rotation rates (1000 Hz and 
up), based on the results of \citet{flame_wave1}.}

The {\tt F500} and {\tt F1000} runs both qualitatively resemble the flame structure in 
\citet{flame_wave1} --- a laterally propagating flame that is lifted off of the bottom of the fuel 
layer --- but they differ in their burning structures. Figures~\ref{fig:time_series_500} and 
\ref{fig:time_series_1000} show time series of the mean molecular weight, $\bar{A}$, for the {\tt 
F500} and {\tt F1000} runs.  Compared to those in \cite{flame_wave1}, ashes behind the flame do not 
reach as high atomic weights. This is not surprising, since those previous runs artificially 
boosted the reaction rates.  Comparing these two new runs, the burning is much more evolved for the 
higher rotation rate, and the ash is actually able to move ahead of the flame front (visible in the 
Figure  \ref{fig:time_series_1000} 100 ms snapshot). We believe that this is because the increased 
rotation better confines the initial perturbation and subsequent expansion from the burning, 
increasing the temperature and density in the flame front such that the reaction rate increases, 
which allows the reactions to progress further. In contrast to {\tt F500} and {\tt F1000}, the 
lowest rotation run --- {\tt F250} --- failed to ignite. The lack of ignition for {\tt F250} also 
aligns with the reasoning given above, with the lower rotation in this case potentially leading to 
insufficient confinement such that the temperature and density required for ignition is not 
achieved. In this scenario, another source of confinement (e.g.\ magnetic fields) would need to 
take over at lower rotation rates to allow a burst to occur, at least for the initial model used 
here.

@@

We found in \citet{flame_wave1} that the flame speed, $s$, obeys the following relation:
\begin{equation}
	\label{eqn:flame_speed_scaling}
	s \appropto L_R (\enucdot)^{\half} \propto \frac{(\enucdot)^{\half}}{\Omega},
\end{equation}
where $\enucdot$ is the specific nuclear energy generation rate and $L_R$ is the Rossby length 
described by Equation \ref{eqn:rossby}. 
This was consistent with the results of \citet{cavecchi:2013}, who noted that the flame speeds in 
their simulations scaled with the Rossby length. As seen in Figure~\ref{fig:flame_speeds_1}, there 
is no obvious inverse scaling of the flame speed with rotation rate in this set of runs. We 
observed earlier, however, that nuclear reactions progress more quickly at higher rotation 
rate. This results in a higher $\enucdot$ --- up to four to six times higher near the burning front 
after the flame ignites --- which may counteract the reduction in flame speed from the increased 
Coriolis confinement. We also observe that $\tt{F500}$ accelerates 
faster than $\tt{F1000}$, which appears to experience a small deceleration at early times. 
\MarginPar{KE: This is speculation on my part -- does someone else have clear idea why?} This 
disparity may be a direct result of the difference in Coriolis force. It could also be due to the 
500 Hz flame taking more time to develop fully and reach higher burning rates, resulting in an 
$\enucdot$ increase over time.

@@

In future simulations, we would like to improve and expand our simulations in order to better understand the processes at play and to include more physics. This includes adding tracer particles to the simulations so we can monitor the fluid motion; extending our simulations to 3D, which would hopefully alleviate some of the boundary effects we have observed in these simulations but will require significant computational resource; and exploring the resolution of our simulations more so that we can ensure that we have resolved all the necessary physical processes. We would also like to model H/He flames, \MarginPar{someone please describe here why these would be interesting as I can't remember what makes them different}. We recently added an MHD solver to \castro~\citep{sazo2020thesis}; this will allow us to explore the effects of magnetic fields on flame propagation in XRBs.
